---
title: "5626 Project 1"
author: "Lisa Siefker"
date: "2025-03-30"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr) 
library(tidyverse)
library(ggplot2) 
```


```{r}
#read data 
gpa_data <- read.csv("GPAAdmissions.csv", header = TRUE)  

head(gpa_data)
```
```{r}
summary(gpa_data)
```
```{r}
str(gpa_data)
```
```{r}
# get counts
gpa_data %>% count(Gender)
gpa_data %>% count(White)
gpa_data %>% count(Asian)
gpa_data %>% count(College)
gpa_data %>% count(Edu_Parent1)
gpa_data %>% count(Edu_Parent1)

```
```{r}
# create a Race category variable
gpa_data1 <- gpa_data %>%
  mutate(Race = case_when(
    White == 1 ~ "White",
    Asian == 1 ~ "Asian",
    TRUE ~ "Minority"
  ))

#head(gpa_data1)

# create scatterplot 
ggplot(gpa_data1, aes(x = SAT.ACT, y = College_GPA, color = Race)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "SAT/ACT Scores vs. College GPA",
       x = "SAT/ACT Score",
       y = "College GPA",
       color = "Race Category") +
  theme_minimal()

```
```{r}
# create faceted scatterplot by race 
ggplot(gpa_data1, aes(x = SAT.ACT, y = College_GPA, color = Race)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "SAT/ACT Scores vs. College GPA",
       x = "SAT/ACT Score",
       y = "College GPA",
       color = "Race Category") +
  theme_minimal() +
  facet_wrap(~ Race)

```
```{r}
# SAT/ACT score by college type

# create faceted scatterplot by college type
ggplot(gpa_data1, aes(x = SAT.ACT, y = College_GPA, color = College)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "SAT/ACT Scores vs. College GPA",
       x = "SAT/ACT Score",
       y = "College GPA",
       color = "College") +
  theme_minimal() +
  facet_wrap(~ College)

```


```{r}
# fit a simple linear regression model 

model_SATACT_College<- lm(College_GPA ~ SAT.ACT, data = gpa_data1)

summary(model_SATACT_College)
```
The model indicates that for every 1 point increase in SAT/ACT score, college GPA increases by about 0.001456. Therefore, for a 100 point increase in SAT/ACT score, college GPA increases by 0.1456. The p-value is very close to zero, indicating that the results are statistically significant. Therefore, I would reject the null hypothesis that SAT/ACT score is not a significant predictor of college GPA. 

The R squared value is 0.1294, which means that the model explains only 12.94% of the variance in college GPA. This means that other factors likely play a large role in explaining variance in college GPA. 

```{r}
plot(model_SATACT_College)
```

The Residuals v. Fitted and Scale-Location plots shows residuals fairly randomly scattered around zero with no clear pattern, indicating linearity and homoscedasticity.

The QQ Plot shows evidence of non-normality, with deviations at tails and potential outliers. Since the sample size is large, non-normality is not a significant problem. 

```{r}
# fit a simple linear regression model 

model_HS_College <- lm(College_GPA ~ HSGPA, data = gpa_data1)

summary(model_HS_College)
```
```{r}
plot(model_HS_College) 
```


```{r}
# convert college to a factor 
gpa_data1$College <- as.factor(gpa_data1$College)

# fit multiple regression model to predict college GPA using SAT/ACT, high school GPA, parental education and College type
model_SATACT_HS_Edu_College <- lm(
  College_GPA ~ SAT.ACT + HSGPA + Edu_Parent1 + Edu_Parent2 + College,
  data = gpa_data1
)

summary(model_SATACT_HS_Edu_College)
```

The R squared of the multiple regression model is 0.319, which is an improvement over the simple linear regression model. However, the model only explains about 32% of variance in college GPA. 

The model shows that high school GPA is the strongest predictor of ollege GPA. For every one unit increase in high school GPA, college GPA increases 0.64. 

SAT/ACT scores have a significant but modest effect on college GPA. A 200 point increase in SAT/ACT score results in a 0.18 increase in college GPA. 

68% variance suggests that other predictors are missing from the model (e.g., study habits or course load)

```{r}
plot(model_SATACT_HS_Edu_College)
```
```{r}
# convert college to a factor 
gpa_data1$College <- as.factor(gpa_data1$College)

# fit multiple regression model to predict college GPA using SAT/ACT, high school GPA
model_HS_SATACT <- lm(
  College_GPA ~ SAT.ACT + HSGPA, 
  data = gpa_data1
)

summary(model_HS_SATACT)
```

```{r}
# use anova to compare SAT.ACT model to SAT.ACT + HSGPA model 
anova(model_SATACT_College, model_HS_SATACT)
```

```{r}
# add interaction term to model that includes SAT.ACT and HSGPA
# convert college to a factor 
gpa_data1$College <- as.factor(gpa_data1$College)

# fit multiple regression model to predict college GPA using SAT/ACT, high school GPA
model_interaction <- lm(
  College_GPA ~ SAT.ACT + HSGPA + SAT.ACT*HSGPA, 
  data = gpa_data1
)

summary(model_interaction)

```
No improvement with the interaction model 
